Reinforcement Learning for Options on Target Volatility Funds

### 2021-05-25 tex/targetvol.tex
paper

To derive the expression for $\omega_t$ we need to assume a generic continuous semi-martingales dynamics under the risk-neutral measure for the underlying securities, so that we can write \Cref{eq:risk_neutral} as 
\begin{equation}
    \frac{dS_t^i}{S_t^i} = \left(r_t - \mu_t^i \right)dt + \nu_t^i \cdot dW_t,
\label{Equity_process}\end{equation}
where $\nu_t$ is an adapted matrix process ensuring the existence of a solution for the SDE and $W_t$ is a $n$-dimensional vector of Brownian motions under $\mathbb{Q}$. Under these assumptions we can derive an expression for $\omega_t$, and we get\footnote{In all formulae the norm for a vector $a$ is defined as $\|a\|\coloneqq\sqrt{a\cdot a}$.}
\begin{equation}
    \omega_t = \frac{\Bar{\sigma}}{\|\alpha_t\cdot \nu_t \|}.
\end{equation}
Hence, putting this last result in the dynamics of $I_t$ we obtain

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[height=6cm,width=7.2cm]{Activation_functions_Learning_curve_1x40_network_confront_activation_strategy_EUR_8x5_900dpi.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Value_function_Learning_curve_1x40_beta_confront_varying_strategy_EUR_8x5_400dpi_compressed.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Hidden_Layers_Learning_curve_3asset_varystrategy_window1000000_EUR_5e8timestep_comparison_900dpi_compressed.pdf}
	\caption{Fine tuning simulations: on the top-left learning curves with different activation functions, on the top-right grid search on the policy-value trade-off and on bottom-centre learning curves on ANNs architectures. On the horizontal axis the number of training episodes. The solid lines are the moving average of the realized rewards on the last $5\times 10^5$ episodes. The shadows represent the $98\%$ confidence intervals. The dash-dotted horizontal red lines are the conservative TVO prices computed according to the optimal risky portfolio composition through \cref{eq:bs_closed_solution}.}
	\label{fig:fine_tuning}
\end{figure}  
\subsection{Local Volatility Dynamics}
\begin{comment}

@@

>>>>

To derive the expression for $\omega_t$ we need to assume a generic continuous semi-martingales dynamics under the risk-neutral measure for the underlying securities, so that we can write \Cref{eq:risk_neutral} as 
\begin{equation}
    \frac{dS_t^i}{S_t^i} = \left(r_t - \mu_t^i \right)dt + \nu_t^i \cdot dW_t,
\label{eq:Equity_process}\end{equation}
where $\nu_t$ is an adapted matrix process ensuring the existence of a solution for the SDE and $W_t$ is a $n$-dimensional vector of Brownian motions under $\mathbb{Q}$. Under these assumptions we can derive an expression for $\omega_t$, and we get\footnote{In all formulae the norm for a vector $a$ is defined as $\|a\|\coloneqq\sqrt{a\cdot a}$.}
\begin{equation}
    \omega_t = \frac{\Bar{\sigma}}{\|\alpha_t\cdot \nu_t \|}.
\end{equation}
Hence, putting this last result in the dynamics of $I_t$ we obtain

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[height=6cm,width=7.2cm]{Activation_functions_Learning_curve_1x40_network_confront_activation_strategy_EUR_8x5_900dpi.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Value_function_Learning_curve_1x40_beta_confront_varying_strategy_EUR_8x5_400dpi_compressed.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Hidden_Layers_Learning_curve_3asset_varystrategy_window1000000_EUR_5e8timestep_comparison_900dpi_compressed.pdf}
	\caption{Fine tuning simulations: on the top-left learning curves with different activation functions, on the top-right grid search on the policy-value trade-off and on bottom-centre learning curves on ANNs architectures. On the horizontal axis the number of training episodes. The solid lines are the moving average of the realized rewards on the last $5\times 10^5$ episodes. The shadows represent the $98\%$ confidence intervals. The dash-dotted horizontal red lines are the conservative TVO prices computed according to the optimal risky portfolio composition through \cref{eq:bs_closed_solution}.}
	\label{fig:fine_tuning}
\end{figure}  
\subsection{Local Volatility Dynamics}
In this section we study the TVS control problem assuming a local volatility model for the dynamics of the risky assets. Thus in this case we have a diffusive term in \Cref{eq:Equity_process} that is a deterministic function both of time and state, i.e. the spot price, $\nu_t=\nu(t,S_t)$. This additional dependency of volatility makes the problem of finding the optimal strategy non-trivial; in fact if we consider the whole equity smiles then the second order term in the HBJ equation \eqref{eq:HJB_general} is non null and thus a closed formula for $\alpha^*$ is not available anymore. Because of that one must resort on Reinforcement Learning to recover the problem solution. Unlike BS model where the strategy depends only on time, in LV dynamics there are no unnecessary information provided by the state block in \Cref{eq:state} to take the optimal action.

We first apply our RL algorithm in the same market data conditions as in the case showed in \Cref{fig:n_sphere_parameterization} on the right, but in the LV dynamics. This choice is done to compare the optimal solution in two well-known financial models. We keep the same ANN architecture as in previous section: a $5\times8$ FFNN with $\tanh$ activation function. This enables us to perform the training using two different initial guesses for the PPO:
\begin{itemize}
	\item \textbf{Random starting point}: the hidden weights $\theta$ of the ANN are randomly initialized so that the training phase starts from a casual policy.
	\item \textbf{Black \& Scholes initial guess}: the algorithm initialize the values of $\theta$ with those of the ANN trained in the BS environment. 
\end{itemize}    
The idea behind the second guess is that the LV dynamics differs from BS for a corrective term in the HJB equation and thus we guess that a network which has learnt the BS optimal strategy will be a good starting point for the RL algorithm as sub-optimal value. We present in \Cref{fig:local_vol_learning_curves} on the left the resulting learning curves for two ANNs trained with the two initialization methods just described.

\begin{figure}[ht!]
	\centering
	\includegraphics[height=6cm,width=7.2cm]{Learning_curve_all_5x8_localvol_3asset_compressed.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Action_RL_5x8_marketdata_seed80_long.pdf}
	\caption{On the left, learning curve for two $5\times8$ FFNNs with $\tanh$ activation function and $c_1 = 0.7$.  The networks are both trained in the same LV but with different starting point: random policy for the orange line and BS optimal policy for the blue one. The dash-dotted horizontal red line is the conservative BS call option price  computed according to the optimal risky portfolio in \Cref{fig:RL_actions} on the left.
	On the right agent actions in a never-seen MC episode.	
}
	\label{fig:local_vol_learning_curves}
\end{figure}

\noindent Analyzing the training results, we observe that the two curves converges both to the same saturation value which is compatible with the optimal BS price for the TVO curve. The RL agent has learnt in both cases the optimal allocation strategy of the BS model. In fact if we plot the agent actions taken in never-seen scenarios (\Cref{fig:local_vol_learning_curves} on the right), we see that they are in accordance with those of BS.

As told before, in the LV model the optimal strategy is not available \textit{a priori} as in BS and thus we can not say if the solution found by the RL agent is sub-optimal. However we can check if the agent is stuck in a local optimum by building a n\"aif strategy to measure the performance of the RL algorithm; we call this strategy \virgolette{baseline}. Since we have proved in \Cref{subsec:BS_model} that in the BS model the solution is the allocation strategy that maximizes the local drift of the TVS, we choose as baseline a manager that applies path-wise the BS solution \eqref{eq:BS_optimal_strategy}. 
In \Cref{tab:mc_results_lv} we compare the MC prices obtained applying in $10^6$ paths the baseline and the ANNs trained before.
\begin{table}[t]
	\centering
	\begin{tabular}{|c|c|} \hline
		Method & TVO price [EUR]  \\ \hline 
		RL from random $\theta$ & $(1.912 \pm 0.003) \times 10^{-2}$ \\
		RL from BS $\theta$ & $(1.915 \pm 0.003) \times 10^{-2}$ \\
		Baseline & $(1.915 \pm 0.004) \times 10^{-2}$ \\ \hline
	\end{tabular}\caption{MC prices of an atm-spot TVO with LV dynamics for the risky assets and with $K=1\,[\text{EUR}]$, $T = 1\,[\text{yr}]$ and $\bar{\sigma}=5\%$.  The prices are obtained computing three different allocation strategies: the ANN trained starting from a random policy, the ANN trained from the BS strategy and the baseline. }
	\label{tab:mc_results_lv}\end{table}
All the prices computed with different strategies are $2\sigma$-compatible among them. Thus we can assert that in the market data conditions we have considered, the LV optimal strategy coincides with the BS one. Moreover we observe that the baseline strategy and the BS optimal one provide the same TVO price, such as the risky securities dynamics is irrelevant for the resulting price.

Thus we try different configurations for the hedging costs $\mu_t$ and the implied volatility surfaces in order to find a market where the TVO prices  
\begin{comment}

@@

%%Buona simulazione local vol [ ppo2_1.5e8_5x8_3e-4frombaselinelong_displacedmarket_month_2assets_beta0.7_maturity2_seed5343061 ] su macchina remota in TVS_LV-v0
